Machine learning enhanced empirical potentials for metals and alloys

### 2020-03-15 doc/papers/ml-eam/manuscript.tex
ML-EAM revision 2020 Feb.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% 
% Abstract
%
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\begin{abstract}
Interatomic potential or force-field plays a vital role in simulating materials. 
Empirical potentials like the embedded atom method (EAM) and its variant 
angular-dependent potential (ADP) have proven successful in modeling many 
metals qualitatively. In the past few years, machine learning has became the 
key approach for modeling interatomic interactions. Powered by big data and 
efficient optimizers, machine learning interatomic potentials can generally 
approximate to the first-principles accuracy. In this works, we successfully 
integrated machine learning with EAM and ADP. Machine learning can significantly 
boost their performances. For energy and force predictions, machine leaned EAM 
and ADP can be almost as accurate as the spectral neighbor analysis potential 
(SNAP) on the fcc Ni, bcc Mo and Mo-Ni alloy systems, while the molecular 
dynamics simulations performed by the later approach is nealy 1000 times 
slower. Machine learned EAM and ADP can also give first-principle accuracy key 
material properties, such as elastic constants, melting temperatures and surface 
energies. Our results suggest a new and systematic route for developing machine 
learning interatomic potentials. 
\end{abstract}

@@

Atomistic modeling plays a vital role in materials science. \textit{ab initio} 
calculation or force-field based molecular dynamics simulation (MD) are 
effective ways to study, understand or predict chemical and physical properties 
of materials. \textit{ab initio} approaches are generally much more precise but 
they are rarely used on large-scale metallic systems due to their extremely-high 
computation expenses. Physical model based empirical potential (force-field), 
such as the embedded-atom method (EAM) \cite{EAM0,EAM1,EAM2,EAM3,EAM4}, 
modified embedded-atom method (MEAM) \cite{MEAM0,MEAM1,MEAM2,MEAM_Jelinek}, 
bond-order potential (BoP) \cite{BOP0,BOP1,BOP2,BOP3}, or angular-dependent 
potential (ADP) \cite{ADP0,ADP1,ADP2,ADP3,ADP_FeNi}, still plays the major role 
in long-time simulations and these empirical methods can achieve reasonable 
accuracy with much lower computation costs. Empirical potentials 
generally have very few learnable parameters and both microscopic observables 
(energy, forces, virial, etc.) and macroscopic observables (melting 
point, surface energy, etc.) can be used to tune these parameters. Finding 
optimal parameters of empirical potentials is always a challenging task. Global 
optimization (GO) approaches (Basin-Hopping \cite{BH0,BH1}, genetic algorithm 
\cite{PCIPY,PGA}, etc) are traditionally used to find the best possible 
parameters. However, the gradients of the losses with respect to model 
parameters are difficult or even impossible to calcualte. Hence, 
GO optimizations are generally not that effective.

@@

However, MLIPs also have challenges. The large parameter space and lack of 
physical background makes the "big data" a necessarity. The cost of dataset is 
non-negligible. Besides, even "big data" can only cover a small portion of real 
physical environments (temperature, external pressure, etc). Outside the 
training zone, the performances of MILPs may not that stable. For long-time 
molecular dynamics (MD) simulations of large-scale ($10^5$ or more) systems, 
computation efficiency also becomes a major concern. Recent benchmark 
tests \cite{SnapBenchmark} suggest that MILPs are still too expensive. 
At present, most MLIPs are used to examine small to medium ($10^3$ to $10^4$) 
systems \cite{SNAP,SFApp0,SFApp1,SFApp2,soap3}.

@@

The adoption of physical constraints in optimizing MLIPs can be very helpful. 
Without any physical constraint, adding high-quality data points is, perhaps, 
the only effective way to tune MLIPs for specific purposes. On the contrast, 
with physical constraints very few training samples are sufficient.
As an example, the elementary SNAP potentials (Mo, Ni) were fitted with just 
hundreds of structures but they performed quite well on elastic constants and 
surface energies, as in the outer loop \cite{SNAP_Mo} of the fitting process 
such material properties were already used. One must also note that data 
consistency is important. Physical constrains are just supplementary roles in 
developing MLIPs. So the values of the are preferred to be measured with 
first-principle calculations.

@@

The implementation of the machine learned embedded atom mathod and the physical
constraints are beyond the scope of this work. The details will be discussed
in another paper. Here, we will give a brief description.

@@

Figure \ref{fig:Mo_eam_adp} compares the $\rho(r)$, $F(\rho)$, $\phi(r)$, 
$\mu(r)$ and $\omega(r)$ functions. The differences between ML-EAM and ML-ADP 
$\rho(r)$, $\phi(r)$ and $F(\rho)$ are quite small. For ML-ADP, both the dipole 
and the quadrupole terms converge to zero at $\sim 4$ \AA. The dipole term gives
the major contribution.

@@

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% 
% Figure: Mo - EOS
%
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\begin{figure}
\centering
\includegraphics[scale=0.65]{figures/Mo_all_eos.png}
\caption{\label{fig:Mo_eos} The energy-volume curves of the original EAM, 
ML-EAM and DFT for the bcc Mo.}
\end{figure}

@@

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% 
% Figure: Alloy EOS - Plot
%
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\begin{figure*}
\centering
\includegraphics[scale=0.5]{figures/alloy_eos.png}
\caption{\label{fig:alloy_eos} Energy-volume curves of Ni$_3$Mo and Ni$_4$Mo 
obtained with the original Zjw04 EAM, ML-EAM, ML-ADP and DFT.}
\end{figure*}

@@

MBE is one of the most widely used schemes for fitting potential energy 
surfaces. In the many-body expansion scheme, the total energy of a system with N 
atoms can be expressed as the sum of all k-body terms where $k \le N$:
\begin{equation}
\label{eq:many_body_expansion}
E = 
\sum_{i}^{\mathrm{C^N_1}}{E^{(1)}_{i}} +
\sum_{i,j}^{\mathrm{C^N_2}}{E^{(2)}_{ij}} + 
\sum_{i,j,k}^{\mathrm{C^N_3}}{E^{(3)}_{ijk}} + \cdots 
\end{equation}
where $\mathrm{C^N_k}$ is the binomial coefficient and $E^{(k)}$ represents the 
k-body contribution. In many cases, higher-order terms like $E_{ij}$ or 
$E_{ijk}$ are symmetric: $E_{ij}=E_{ji}$, $E_{ijk}=E_{ikj}=E_{jik}=\cdots$. 
Thus, Equation \ref{eq:many_body_expansion} can be further transformed to:
\begin{align}
\label{eq:MBE_atomic}
E & = \sum_{i}^{\mathrm{N}}{\left(
    E^{(1)}_{i} + 
    \frac{1}{2!}\sum_{j \ne i}^{\mathrm{N}}{E^{(2)}_{ij}} +
    \frac{1}{3!}\sum_{j \ne i}^{\mathrm{N}}{
        \sum_{k \ne i,j}^{\mathrm{N}}{E^{(3)}_{ijk}}
    } +
    \cdots
\right)} \nonumber \\
& = \sum_{i}^{\mathrm{N}}{E_{i}}
\end{align}
where $E_{i}$ is the atomic energy of atom $i$.

>>>>

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% 
% Abstract
%
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\begin{abstract}
Interatomic potential or force-field plays a vital role in simulating materials. 
Empirical potentials like the embedded atom method (EAM) and its variant 
angular-dependent potential (ADP) have proven successful in modeling many 
metals qualitatively. In the past few years, machine learning has became the 
key approach for modeling interatomic interactions. Powered by big data and 
efficient optimizers, machine learning interatomic potentials can generally 
approximate to the first-principles accuracy. In this works, we successfully 
integrated machine learning interatomic potential with EAM and ADP. Machine 
learning can significantly boost their performances. For energy and force 
predictions, machine leaned EAM and ADP can be almost as accurate as the 
spectral neighbor analysis potential (SNAP) on the fcc Ni, bcc Mo and Mo-Ni 
alloy systems, while the molecular dynamics simulations performed by the later 
approach is nealy 1000 times slower. Machine learned EAM and ADP can also give 
first-principle accuracy key material properties, such as elastic constants, 
melting temperatures and surface energies. Our results suggest a new and 
systematic route for developing machine learning interatomic potentials. 
\end{abstract}

@@

Atomistic modeling plays a vital role in materials science. \textit{Ab initio} 
calculation or force-field based molecular dynamics simulation (MD) are 
effective ways to study, understand or predict chemical and physical properties 
of materials. \textit{ab initio} approaches are generally much more precise but 
they are rarely used on large-scale metallic systems due to their extremely-high 
computation expenses. Physical model based empirical potential (force-field), 
such as the embedded-atom method (EAM) \cite{EAM0,EAM1,EAM2,EAM3,EAM4}, 
modified embedded-atom method (MEAM) \cite{MEAM0,MEAM1,MEAM2,MEAM_Jelinek}, 
bond-order potential (BoP) \cite{BOP0,BOP1,BOP2,BOP3}, or angular-dependent 
potential (ADP) \cite{ADP0,ADP1,ADP2,ADP3,ADP_FeNi}, still plays the major role 
in long-time simulations and these empirical methods can achieve reasonable 
accuracy with much lower computation costs. Empirical potentials 
generally have very few learnable parameters and both microscopic observables 
(energy, forces, virial, etc.) and macroscopic observables (melting 
point, surface energy, etc.) can be used to tune these parameters. Finding 
optimal parameters of empirical potentials is always a challenging task. Global 
optimization (GO) approaches (Basin-Hopping \cite{BH0,BH1}, genetic algorithm 
\cite{PCIPY,PGA}, etc) are traditionally used to find the best possible 
parameters. However, the gradients of the losses with respect to model 
parameters are difficult or even impossible to calcualte. Hence, 
global optimizations are generally not that effective.

@@

However, MLIPs also have challenges. The large parameter space and lack of 
physical background makes the "big data" a requisite. The cost of dataset is 
non-negligible. Besides, even "big data" can only cover a small portion of real 
physical environments (temperature, external pressure, etc). Outside the 
training zone, the performances of MILPs may not that stable. For long-time 
molecular dynamics (MD) simulations of large-scale ($10^5$ or more) systems, 
computation efficiency also becomes a major concern. Recent benchmark 
tests \cite{SnapBenchmark} suggest that MILPs are still too expensive. 
At present, most MLIPs are used to examine small to medium ($10^3$ to $10^4$) 
systems \cite{SNAP,SFApp0,SFApp1,SFApp2,soap3}.

@@

The adoption of physical constraints in optimizing MLIPs can be very helpful. 
Without any physical constraint, adding high-quality data points is, perhaps, 
the only effective way to tune MLIPs for specific purposes. On the contrast, 
with physical constraints very few training samples are sufficient.
As an example, the elementary SNAP potentials (Mo, Ni) were fitted with just 
hundreds of structures but they performed quite well on elastic constants and 
surface energies, as in the outer loop \cite{SNAP_Mo} of the fitting process 
such material properties were already used. One must also note that data 
consistency is important. Physical constraints are just supplementary roles in 
developing MLIPs, so they are preferred to be measured with first-principle 
calculations.

@@

The implementation of the machine learned EAM and the physical
constraints are beyond the scope of this work. The details will be discussed
in another paper. Here, we will give a brief description.

@@

Figure \ref{fig:Ni_eam} compares the $\rho(r)$, $F(\rho)$, $\phi(r)$, 
$\mu(r)$ and $\omega(r)$ functions. The differences between ML-EAM and ML-ADP 
$\rho(r)$, $\phi(r)$ and $F(\rho)$ are quite small. For ML-ADP, both the dipole 
and the quadrupole terms converge to zero at $\sim 4$ \AA. The dipole term gives
the major contribution.

@@

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% 
% Figure: Mo - EOS
%
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\begin{figure}
\centering
\includegraphics[scale=0.65]{figures/Mo_all_eos.png}
\caption{\label{fig:Mo_eos} The energy-volume curves of the original EAM, 
ML-EAM, ML-ADP and DFT for the bcc Mo.}
\end{figure}

@@

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
% 
% Figure: Alloy EOS - Plot
%
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
\begin{figure*}
\centering
\includegraphics[scale=0.5]{figures/alloy_eos.png}
\caption{\label{fig:alloy_eos} Energy-volume curves of Ni$_3$Mo and Ni$_4$Mo 
obtained with the original Zjw04 EAM, ML-EAM, ML-ADP, SNAP and DFT.}
\end{figure*}

@@

MBE is one of the most widely used schemes for fitting potential energy 
surfaces. In the many-body expansion scheme, the total energy of a system with N 
atoms can be expressed as the sum of all k-body terms where $k \le N$:
\begin{equation}
\label{eq:many_body_expansion}
E = 
\sum_{i}^{\mathrm{C^N_1}}{E^{(1)}_{i}} +
\sum_{i,j}^{\mathrm{C^N_2}}{E^{(2)}_{ij}} + 
\sum_{i,j,k}^{\mathrm{C^N_3}}{E^{(3)}_{ijk}} + \cdots 
\end{equation}
where $\mathrm{C^N_k}$ (k=1,2,3) is the binomial coefficient and $E^{(k)}$ 
represents the k-body contribution. In many cases, higher-order terms like 
$E_{ij}$ or 
$E_{ijk}$ are symmetric: $E_{ij}=E_{ji}$, $E_{ijk}=E_{ikj}=E_{jik}=\cdots$. 
Thus, Equation \ref{eq:many_body_expansion} can be further transformed to:
\begin{align}
\label{eq:MBE_atomic}
E & = \sum_{i}^{\mathrm{N}}{\left(
    E^{(1)}_{i} + 
    \frac{1}{2!}\sum_{j \ne i}^{\mathrm{N}}{E^{(2)}_{ij}} +
    \frac{1}{3!}\sum_{j \ne i}^{\mathrm{N}}{
        \sum_{k \ne i,j}^{\mathrm{N}}{E^{(3)}_{ijk}}
    } +
    \cdots
\right)} \nonumber \\
& = \sum_{i}^{\mathrm{N}}{E_{i}}
\end{align}
where $E_{i}$ is the atomic energy of atom $i$.
